Migration of Gas Giant Planets in Gravitationally Unstable Disks

### 2011-05-30 Text/migration.tex
Dick's Final Changes May 30, 2011
git-svn-id: http://jaga.astro.indiana.edu/paperrepo/ApJL/Migration2010@85 3b7f7548-5c66-0410-83ab-3997167635fc

\begin{abstract}
Characterization of migration in gravitationally unstable disks is necessary to understand the fate of protoplanets formed by disk instability. As part of a larger study, we are using a 3D radiative hydrodynamics code to investigate how an embedded gas giant planet interacts with a gas disk that undergoes gravitational instabilities (GIs). This Letter presents preliminary results from simulations with a Jupiter-mass planet placed in orbit at 25 AU within a 0.14 $M_{\odot}$ disk. The disk spans 5 to 40 AU around a 1 $M_{\odot}$ star and is initially close to marginal stability. In one simulation, the planet is inserted prior to the eruption of GIs; in another, it is inserted only after the disk has settled into a quasi-steady GI-active state, where heating by GIs roughly balances radiative cooling. When the planet is present from the beginning, its own wake stimulates growth of a particular global mode with which it strongly interacts, and the planet plunges inward six AU in about 10$^3$ years. In both cases with embedded planets, there are times when the planet's radial motion is slow and varies in direction. At other times, when the planet appears to be interacting with strong spiral modes, migration both inward and outward can be relatively rapid, covering several AUs over hundreds of years. Migration in both cases appears to stall near the inner Lindblad resonance of a dominant low-order mode. Planet orbit eccentricities fluctuate rapidly between about 0.02 to 0.1 throughout the GI-active phases of the simulations. 
\end{abstract}

@@

Migration of gas giant planets due to interactions with a circumstellar gas disk can play a major role in defining the architecture of planetary systems. Work on migration \citep[see review by][]{papaloizou2007} has included gravitational interaction of planets with both laminar and turbulent disks. However, radiative transport, detailed equations of state (EOS), and the self-gravity of the gas disk have usually been ignored, the effects of a non-isothermal EOS have only recently been included \citep[e.g.,][]{paardekooper2006,paardekooper2010,paardekooper2011}. 
\citet{boss2005} and \citet{mayer2004} examined radial migration of planet-mass fragments in gravitationally unstable disks, but their disks were violently disrupted by fragmentation under conditions (radii $<$ 40 AU, disk-to-star mass ratios $M_d/M_s \sim 0.1$, and stellar mass $M_s \sim 1\;M_{\odot}$) where fragmentation may not actually occur  \citep{rafikov2005,rafikov2007,boley2006,boley2007b,boley2008,forgan2009,cai2010}. More recently, fragmentation into clumps with gas giant or brown dwarf masses has been documented in numerical simulations of disks that are relatively massive ($M_d/M_s \sim$ a few tenths) and spatially extended (outer radii $>$ 50 AU) \citep{krumholz2007,stamatellos2007,stamatellos2009,boley2009,boley2010}, where fragmentation is expected from semi-analytic arguments \citep[e.g.,][]{clarke2009,rafikov2009,dodson2009}. The fate of the clumps then depends in part on their radial migration, which is a chaotic and messy affair in a fragmenting disk \citep[e.g.,][]{boley2009,boley2010,vorobyov2010,boley2010b}. The occurrence of gravitational instabilities (GIs) may also be episodic \citep[e.g.,][]{vorobyov2006,vorobyov2010b,zhu2010}. Clumps that survive and contract to the dimensions of young planets can later find themselves in a disk that erupts again into GI activity. As the star/disk system evolves, such a protoplanet may end up in a region of a GI-active disk where fragmentation does not occur.

@@

This Letter presents three simulations. The first, which we call the {\sl fiducial run}, goes from time $t = 0$ to 21 ORP without a planet. In the second simulation, which we call the {\sl t = 0 planet run}, a one Jupiter-mass ($M_J$) planet with a roughly circular midplane orbit is included in the initial $t = 0$ ORP equilibrium disk at 25 AU. In the third simulation, which we call the {\sl t = 10 planet run}, a 1 $M_J$ planet with a similar orbit is inserted into the fiducial run at $t = 10.5$ ORP. These three simulations are part of a much larger suite, to be described elsewhere, in which the planet mass is varied. We estimate crude semi-major axes $a$ and eccentricities $e$ from the planet's $\varpi(t)$ by using the maximum and minimum radii, $\varpi_{\rm max}$ and $\varpi_{\rm min}$, attained by the planet over each interval that the planet's $\phi$-coordinate changes by 2$\pi$ (one ``orbit'').  Then, for that orbit, $a = (\varpi_{\rm max}+\varpi_{\rm min})/2$ and $e = (\varpi_{\rm max}-\varpi_{\rm min})/(\varpi_{\rm max}+\varpi_{\rm min})$. These parameters are only indicative; there are strong orbit to orbit variations in $a$ and $e$.

@@

{\it The $t = 0$ Planet Run.} Figure \ref{fig:Am} shows that inclusion of even this modest mass planet ($0.7 \%$ of $M_d$) has a dramatic effect on the burst phase. Without a planet, when all modes grow from imposed noise, it takes 4 ORPs for coherent spiral modes to organize and grow to nonlinear amplitude. The growth is centered near the $Q$-minimum at $26$ AU. Modes of cos($m\phi$) symmetry with $m = 4$ and 5 dominate. An $m = 3$ (three-armed) spiral also grows, but somewhat more slowly, and lags the other modes in time. In the $t = 0$ planet run, the planet develops an organized wake within about an ORP in which $m = 3$ is a dominant component ($\sim 5$ \% global density perturbation). This nonlinear seeding causes $m = 3$ to dominate the GI burst, which also occurs $\sim$ 2 ORP earlier. Because of our initial placement of the planet inside but close to the $Q$-minimum, the corotation radius (CR) of the $m = 3$ mode is fairly close to that of the planet's orbit period. When the triggered $m = 3$ mode becomes strongly nonlinear at about 3 ORP, planet migration is significantly affected.

@@

Even in non-self-gravitating disks, the presence of discrete radial structure is known to cause instabilities \citep{li1999,li2000}. \citet{meschiari2008} demonstrated that a gap, presumably opened by an embedded planet, can cause global modes to become unstable in self-gravitating disks that are otherwise stable against the development of GIs. Here we are dealing with a planet that is not quite massive enough to open a gap, but its gravitational interaction with the disk also has a strong effect on the growth and onset of GIs.

Figure \ref{fig:a} suggests that the overall radial drift during the first 10 ORP of the $t = 0$ planet run is not too different from what one would get by extrapolating the migration due solely to the planet's wake during the first 2 or 3 ORP, where the average radial migration is about 4$\times10^{-3}$ AU yr$^{-1}$. Although there are no simple formulae that are fully applicable to our radiatively cooled disk, using eq. (70) from \citet{tanaka2002} with parameters from the simulated disk, we estimate 6$\times10^{-3}$ AU yr$^{-1}$ as an analytic expectation. This is probably a reasonable estimate for comparison purposes, because our resolution is probably insufficient to compute the corotation torque \citep{paardekooper2010}. 
While the interaction with a GI-active disk does not seem to result in a drastically different overall migration rate, the direction and magnitude of migration in both simulations is quite variable and seems to depend strongly on interactions of the planet with discrete modes in the disk. In $t = 10$ planet run, the overall radial migration rate is similar but is even more erratic in direction than in the $t = 0$ planet run. The important departures from laminar disk theory are the more erratic nature of migration, the importance of interactions with discrete low-order modes, and stalling of migration near the inner edge of the strongest GI activity. The migration is neither a random walk nor monotonic but has a chaotic character due to nonlinear interactions. For all these reasons, it will be important in future studies to explore the dependence of the planet's behavior on its mass and on its placement relative to the site of GI-eruptions (for $t = 0$ type runs) and relative to the CRs of low-order GI modes (for $t = 10$ type runs).

In both simulations, the inward migration appears to halt near 17 AU. This may simply be a coincidence, because the overall average migration rates are similar, the starting positions are essentially the same, and the calculations have a similar duration. We plan to extend simulations like these to much longer times in the future. However, the final radial positions of both planets happen to be the location of the Inner Lindblad resonance (ILR) for a strong m = 2 mode present in both runs. This ILR is associated with a surface density enhancement in the disk. Near this enhancement, the surface density gradient may be falling steeply enough to halt migration, a mechanism that can work in a laminar disk \citep{paardekooper2009}. 
Alternatively, at 17 AU, the level of GI activity is quite different interior to the planet ($Q \gtrsim 2$) and exterior to the planet ($Q \lesssim 2$), and this may affect the balance of the Lindblad torques exerted by the disk on the planet. Whether the migration really does stall and what the mechanism may be are subjects for further research.

>>>>

\begin{abstract}
Characterization of migration in gravitationally unstable disks is necessary to understand the fate of protoplanets formed by disk instability. As part of a larger study, we are using a 3D radiative hydrodynamics code to investigate how an embedded gas giant planet interacts with a gas disk that undergoes gravitational instabilities (GIs). This Letter presents preliminary results from simulations with a Jupiter-mass planet placed in orbit at 25 AU within a 0.14 $M_{\odot}$ disk. The disk spans 5 to 40 AU around a 1 $M_{\odot}$ star and is initially marginally unstable. In one simulation, the planet is inserted prior to the eruption of GIs; in another, it is inserted only after the disk has settled into a quasi-steady GI-active state, where heating by GIs roughly balances radiative cooling. When the planet is present from the beginning, its own wake stimulates growth of a particular global mode with which it strongly interacts, and the planet plunges inward six AU in about 10$^3$ years. In both cases with embedded planets, there are times when the planet's radial motion is slow and varies in direction. At other times, when the planet appears to be interacting with strong spiral modes, migration both inward and outward can be relatively rapid, covering several AUs over hundreds of years. Migration in both cases appears to stall near the inner Lindblad resonance of a dominant low-order mode. Planet orbit eccentricities fluctuate rapidly between about 0.02 to 0.1 throughout the GI-active phases of the simulations. 
\end{abstract}

@@

Migration of gas giant planets due to interactions with a circumstellar gas disk can play a major role in defining the architecture of planetary systems. Work on migration \citep[see review by][]{papaloizou2007} has included gravitational interaction of planets with both laminar and turbulent disks. However, radiative transport, detailed equations of state (EOS), and the self-gravity of the gas disk have usually been ignored; the effects of a non-isothermal EOS have only recently been included \citep[e.g.,][]{paardekooper2006,paardekooper2010,paardekooper2011}. 
\citet{boss2005} and \citet{mayer2004} examined radial migration of planet-mass fragments in gravitationally unstable disks, but their disks were violently disrupted by fragmentation under conditions (radii $<$ 40 AU, disk-to-star mass ratios $M_d/M_s \sim 0.1$, and stellar mass $M_s \sim 1\;M_{\odot}$) where fragmentation may not actually occur  \citep{rafikov2005,rafikov2007,boley2006,boley2007b,boley2008,forgan2009,cai2010}. More recently, fragmentation into clumps with gas giant or brown dwarf masses has been documented in numerical simulations of disks that are relatively massive ($M_d/M_s \sim$ a few tenths) and spatially extended (outer radii $>$ 50 AU) \citep{krumholz2007,stamatellos2007,stamatellos2009,boley2009,boley2010}, where fragmentation is expected from semi-analytic arguments \citep[e.g.,][]{clarke2009,rafikov2009,dodson2009}. The fate of the clumps then depends in part on their radial migration, which is a chaotic and messy affair in a fragmenting disk \citep[e.g.,][]{boley2009,boley2010,vorobyov2010,boley2010b}. The occurrence of gravitational instabilities (GIs) may be episodic \citep[e.g.,][]{vorobyov2006,vorobyov2010b,zhu2010}. Clumps that survive and contract to the dimensions of young planets can later find themselves in a disk that erupts again into GI activity. As the star/disk system evolves, such a protoplanet may end up in a region of a GI-active disk where fragmentation does not occur.

@@

This Letter presents three simulations. The first, which we call the {\sl fiducial run}, goes from time $t = 0$ to 21 ORP without a planet. In the second simulation, which we call the {\sl t = 0 planet run}, a one Jupiter-mass ($M_J$) planet with a roughly circular midplane orbit is included in the initial $t = 0$ ORP equilibrium disk at 25 AU. In the third simulation, which we call the {\sl t = 10 planet run}, a 1 $M_J$ planet with a similar orbit is inserted into the fiducial run at $t = 10.5$ ORP. These three simulations are part of a larger suite, to be described elsewhere, in which the planet mass is varied. We estimate crude semi-major axes $a$ and eccentricities $e$ from the planet's $\varpi(t)$ by using the maximum and minimum radii, $\varpi_{\rm max}$ and $\varpi_{\rm min}$, attained by the planet over each interval that the planet's $\phi$-coordinate changes by 2$\pi$ (one ``orbit'').  Then, for that orbit, $a = (\varpi_{\rm max}+\varpi_{\rm min})/2$ and $e = (\varpi_{\rm max}-\varpi_{\rm min})/(\varpi_{\rm max}+\varpi_{\rm min})$. These parameters are only indicative; there are strong orbit to orbit variations in $a$ and $e$.

@@

{\it The $t = 0$ Planet Run.} Figure \ref{fig:Am} shows that inclusion of even this modest mass planet ($0.7 \%$ of $M_d$) has a dramatic effect on the burst phase. Without a planet, when all modes grow from imposed noise, it takes 4 ORPs for coherent spiral modes to organize and grow to nonlinear amplitude. The growth is centered near the $Q$-minimum at $26$ AU. Modes of cos($m\phi$) symmetry with $m = 4$ and 5 dominate. An $m = 3$ (three-armed) spiral also grows, but somewhat more slowly, and lags the other modes in time. In the $t = 0$ planet run, the planet develops an organized wake within about an ORP in which $m = 3$ is a dominant component ($\sim 5$ \% global density perturbation). This nonlinear seeding causes $m = 3$ to dominate the GI burst, which also occurs $\sim$ 2 ORP earlier. Because of our initial placement of the planet inside but close to the $Q$-minimum, the corotation radius (CR) of the $m = 3$ mode is fairly close to that of the planet's orbit radius. When the triggered $m = 3$ mode becomes strongly nonlinear at about 3 ORP, planet migration is significantly affected.

@@

Even in non-self-gravitating disks, the presence of discrete radial structure is known to cause instabilities \citep{li2000,li2001}. \citet{meschiari2008} demonstrated that a gap, presumably opened by an embedded planet, can cause global modes to become unstable in self-gravitating disks that are otherwise stable against the development of GIs. Here we are dealing with a planet that is not quite massive enough to open a gap, but its gravitational interaction with the disk also has a strong effect on the growth and onset of GIs.

Figure \ref{fig:a} suggests that the overall radial drift during the first 10 ORP of the $t = 0$ planet run is not too different from what one would get by extrapolating the migration due solely to the planet's wake during the first 2 or 3 ORP, where the average radial migration is about 4$\times10^{-3}$ AU yr$^{-1}$. Although there are no simple formulae that are fully applicable to our radiatively cooled disk, using eq. (70) from \citet{tanaka2002} with parameters from the simulated disk, we estimate 6$\times10^{-3}$ AU yr$^{-1}$ as an analytic expectation. This is a reasonable estimate for comparison purposes, because our resolution is probably insufficient to compute the corotation torque \citep{paardekooper2010}. 
While the interaction with a GI-active disk does not seem to result in a drastically different overall migration rate, the direction and magnitude of migration in both simulations is quite variable and seems to depend strongly on interactions of the planet with discrete modes in the disk. In $t = 10$ planet run, the overall radial migration rate is similar but is even more erratic in direction than in the $t = 0$ planet run. The important departures from laminar disk theory are threefold: the more erratic nature of migration, the importance of interactions with discrete low-order modes, and stalling of migration near the inner edge of the strongest GI activity. The migration is neither a random walk nor monotonic but has a chaotic character due to nonlinear interactions. For all these reasons, it will be important in future studies to explore the dependence of the planet's behavior on its mass and on its placement relative to the site of GI-eruptions (for $t = 0$ type runs) and relative to the CRs of low-order GI modes (for $t = 10$ type runs).

In both simulations, the inward migration appears to halt near 17 AU. This may simply be a coincidence, because the overall average migration rates are similar, the starting positions are essentially the same, and the calculations have a similar duration. We plan to extend simulations like these to much longer times in the future. However, the final radial positions of both planets happen to be the location of the inner Lindblad resonance (ILR) for a strong m = 2 mode present in both runs. This ILR is associated with a surface density enhancement in the disk. Near this enhancement, the surface density gradient may be falling steeply enough to halt migration, a mechanism that can work in a laminar disk \citep{paardekooper2009}. 
Alternatively, at 17 AU, the level of GI activity is quite different interior to the planet ($Q \gtrsim 2$) and exterior to the planet ($Q \lesssim 2$), and this may affect the balance of the Lindblad torques exerted by the disk on the planet. Whether the migration really does stall and what the mechanism may be are subjects for further research.
